Trap spaces of Boolean networks are conflict-free siphons of their Petri net encoding

### 2023-02-24 manuscripts/TCS2022/tcs2022.tex
more plots

\usepackage{pgfplots}
\pgfplotsset{width=10cm,compat=1.13}

@@

We here showcase a theoretical application of the connection between trap spaces in Boolean networks and conflict-free siphons in Petri nets.
We use it to prove a property of minimal trap spaces, which has surprisingly not been formally proved.
Specifically, all minimal trap spaces of a Boolean network are mutually disjoint.
This property is important because we can use it to approximate the set of attractors of the Boolean network~\cite{klarner2015computing}.

@@

\setlength{\tabcolsep}{4pt}
\begin{table}[!htb]
  \caption{Results on the real-world models from the BBM repository.}
  \centering%
  \label{tab:bbm_repo}
  \begin{tabular}{lrrrr}
    \toprule
    Method & \# failures & avg-lqm (s) & avg-mono (s) & avg-all (s) \\ \midrule
    \texttt{bioLQM} & 9 (134) & 12.87 & N/A & N/A \\
    \texttt{PyBoolNet} & 12 & 8.87 & 11.00 & 13.59 \\
    \texttt{mpbn} & 2 (187) & N/A & 2.31 & N/A \\
    \texttt{Trappist}-MaxSAT & 1 & 0.03 & 1.09 & 1.01 \\
    \texttt{Trappist}-CP & - & - & - & - \\
    \texttt{Trappist}-ILP & - & - & - & - \\
    \texttt{Trappist}-ASP & 1 & 0.05 & 1.02 & 0.93 \\
    \bottomrule
  \end{tabular}
\end{table}
\setcounter{rownumber}{0}

@@

	\addplot[
		smooth,
		thick,
		%dashed,
		color=magenta,
		mark=square,
	]
	coordinates {
		(0.5, 164)(1, 190)(2, 214)(5, 249)(10, 266)(20, 291)(50, 317)(100, 328)(180, 338)
	}; \addlegendentry{\texttt{Trappist-maxSAT}}

	\addplot[
		smooth,
		thick,
		dashed,
		color=violet,
		mark=star,
	]
	coordinates {
		(0.5, 179)(1, 211)(2, 244)(5, 274)(10, 304)(20, 327)(50, 339)(100, 347)(180, 349)
	}; \addlegendentry{\texttt{Trappist-ASP}}

	\end{semilogxaxis}
	\end{tikzpicture}
	\caption{Cumulative numbers of random models solved by the competing methods with respect to enumerating the first 1000 minimal trap spaces.}
	\label{fig:random_cumul_first}
\end{figure}

@@

Table~\ref{tab:result_N_K} shows the experimental results on N-K models.
Column \(n\) denotes the network size.
Columns \texttt{LQM} and \texttt{PBN} show the results of \texttt{bioLQM} and \texttt{PyBoolNet}, respectively.
For each method, the number outside the parentheses indicates the number of failures, whereas the number inside the parentheses indicates the average running time (in seconds).
Note that when computing the average running time, if the running time exceeds 180s, it is considered as 180s.
From these results, we obtained several observations consistent with those obtained for real-world models.

TODO: \dots

@@

Second, surprisingly \texttt{bioLQM} could not handle any model.
One of the reason may be that the BDD characterizing all trap spaces is too large, and its computation is slow.
It is apparent because the network size is large (\(\geq 100\)) and the Boolean functions are not simple.

>>>>

\usepackage{pgfplots}
\pgfplotsset{width=10cm,compat=1.13}
\usepgfplotslibrary{groupplots}

@@

We here showcase a theoretical application of the connection between trap spaces in Boolean networks and conflict-free siphons in Petri nets.
We use it to prove a property of minimal trap spaces, which has surprisingly not been formally proved.
Specifically, all minimal trap spaces of a Boolean network are mutually disjoint.
This property is important because we can use it to approximate the set of attractors of the Boolean network under any update scheme~\cite{klarner2015computing} or to compute exactly the set of complex attractors of the Boolean network under the fully asynchronous update scheme~\cite{DBLP:conf/bcb/TrinhHB22}.

@@

%\setlength{\tabcolsep}{4pt}
%\begin{table}[!htb]
%  \caption{Results on the real-world models from the BBM repository.}
%  \centering%
%  \label{tab:bbm_repo}
%  \begin{tabular}{lrrrr}
%    \toprule
%    Method & \# failures & avg-lqm (s) & avg-mono (s) & avg-all (s) \\ \midrule
%    \texttt{bioLQM} & 9 (134) & 12.87 & N/A & N/A \\
%    \texttt{PyBoolNet} & 12 & 8.87 & 11.00 & 13.59 \\
%    \texttt{mpbn} & 2 (187) & N/A & 2.31 & N/A \\
%    \texttt{Trappist}-MaxSAT & 1 & 0.03 & 1.09 & 1.01 \\
%    \texttt{Trappist}-CP & - & - & - & - \\
%    \texttt{Trappist}-ILP & - & - & - & - \\
%    \texttt{Trappist}-ASP & 1 & 0.05 & 1.02 & 0.93 \\
%    \bottomrule
%  \end{tabular}
%\end{table}
%\setcounter{rownumber}{0}

\begin{figure}[ht]
	\centering
	\begin{tikzpicture}
		\begin{semilogxaxis}[
			%title={},
			xlabel={Time (s)},
			ylabel={Models solved},
			xmin=-1, xmax=180,
			log basis x={10},
			%scale only axis,
			ymin=100, ymax=134,
			%ymode=log,
			%log basis y={10},
			xtick={0.5, 1, 2, 5, 10, 20, 50, 100, 180},
			xticklabels = {0.5, 1, 2, 5, 10, 20, 50, 100, 180},
			ytick={100, 110, 120, 130},
			y=0.5cm/4,
			%legend style={font=\fontsize{7}{8}\selectfont},
			%legend style={at={(1.2,-0.1)},anchor=north west},
			legend pos=outer north east,
			ymajorgrids=true,
			grid style=dashed,
		]

		\addplot[
			smooth,
			thick,
			dashed,
			color=orange,
			mark=asterisk,
		]
		coordinates {
			(0.5, 115)(1, 118)(2, 121)(5, 122)(10, 122)(20, 123)(50, 125)(100, 125)(180, 125)
		}; \addlegendentry{\texttt{bioLQM}}

		\addplot[
			smooth,
			thick,
			color=blue,
			mark=*,
		]
		coordinates {
			(0.5, 108)(1, 118)(2, 121)(5, 126)(10, 126)(20, 126)(50, 128)(100, 128)(180, 128)
		}; \addlegendentry{\texttt{PyBoolNet}}

		\addplot[
			smooth,
			thick,
			%dashed,
			color=cyan,
			mark=triangle,
		]
		coordinates {
			(0.5, 116)(1, 117)(2, 117)(5, 117)(10, 117)(20, 117)(50, 118)(100, 118)(180, 118)
		}; \addlegendentry{\texttt{mpbn}}

		\addplot[
			smooth,
			thick,
			%dashed,
			color=teal,
			mark=square,
		]
		coordinates {
			(0.5, 133)(1, 134)(2, 134)(5, 134)(10, 134)(20, 134)(50, 134)(100, 134)(180, 134)
		}; \addlegendentry{\texttt{maxSAT}}

%		\addplot[
%			smooth,
%			thick,
%			%dashed,
%			color=green,
%			mark=star,
%		]
%		coordinates {
%			(0.5, 143)(1, 176)(2, 195)(5, 222)(10, 249)(20, 264)(50, 287)(100, 308)(180, 322)
%		}; \addlegendentry{\texttt{CP}}

%		\addplot[
%			smooth,
%			thick,
%			dashed,
%			color=magenta,
%			mark=star,
%		]
%		coordinates {
%			(0.5, 144)(1, 177)(2, 196)(5, 223)(10, 250)(20, 265)(50, 288)(100, 309)(180, 323)
%		}; \addlegendentry{\texttt{ILP}}

		\addplot[
			smooth,
			thick,
			dashed,
			color=violet,
			mark=star,
		]
		coordinates {
			(0.5, 131)(1, 134)(2, 134)(5, 134)(10, 134)(20, 134)(50, 134)(100, 134)(180, 134)
		}; \addlegendentry{\texttt{ASP}}

		\end{semilogxaxis}
	\end{tikzpicture}
	\begin{tikzpicture}
		\begin{semilogxaxis}[
			%title={},
			xlabel={Time (s)},
			ylabel={Models solved},
			xmin=-1, xmax=180,
			log basis x={10},
			%scale only axis,
			ymin=150, ymax=211,
			%ymode=log,
			%log basis y={10},
			xtick={0.5, 1, 2, 5, 10, 20, 50, 100, 180},
			xticklabels = {0.5, 1, 2, 5, 10, 20, 50, 100, 180},
			ytick={150, 170, 190, 210},
			y=0.5cm/7,
			%legend style={font=\fontsize{7}{8}\selectfont},
			%legend style={at={(0.0,-0.1)},anchor=north},
			legend pos=outer north east,
			ymajorgrids=true,
			grid style=dashed,
		]

		\addplot[
			smooth,
			thick,
			color=blue,
			mark=*,
		]
		coordinates {
			(0.5, 151)(1, 170)(2, 178)(5, 188)(10, 189)(20, 190)(50, 194)(100, 198)(180, 199)
		}; \addlegendentry{\texttt{PyBoolNet}}

		\addplot[
			smooth,
			thick,
			%dashed,
			color=cyan,
			mark=triangle,
		]
		coordinates {
			(0.5, 160)(1, 177)(2, 181)(5, 183)(10, 183)(20, 184)(50, 185)(100, 185)(180, 185)
		}; \addlegendentry{\texttt{mpbn}}

		\addplot[
			smooth,
			thick,
			%dashed,
			color=teal,
			mark=square,
		]
		coordinates {
			(0.5, 191)(1, 204)(2, 208)(5, 210)(10, 210)(20, 210)(50, 210)(100, 210)(180, 210)
		}; \addlegendentry{\texttt{maxSAT}}

		%		\addplot[
		%			smooth,
		%			thick,
		%			%dashed,
		%			color=green,
		%			mark=star,
		%		]
		%		coordinates {
		%			(0.5, 143)(1, 176)(2, 195)(5, 222)(10, 249)(20, 264)(50, 287)(100, 308)(180, 322)
		%		}; \addlegendentry{\texttt{CP}}

		%		\addplot[
		%			smooth,
		%			thick,
		%			dashed,
		%			color=magenta,
		%			mark=star,
		%		]
		%		coordinates {
		%			(0.5, 144)(1, 177)(2, 196)(5, 223)(10, 250)(20, 265)(50, 288)(100, 309)(180, 323)
		%		}; \addlegendentry{\texttt{ILP}}

		\addplot[
			smooth,
			thick,
			dashed,
			color=violet,
			mark=star,
		]
		coordinates {
			(0.5, 205)(1, 208)(2, 209)(5, 210)(10, 210)(20, 210)(50, 210)(100, 210)(180, 210)
		}; \addlegendentry{\texttt{ASP}}

	\end{semilogxaxis}
	\end{tikzpicture}
	\caption{Cumulative numbers of the \texttt{BBM} models that have less than 1000 minimal trap spaces (above) and \texttt{BBM} models (below) solved by the competing methods with respect to enumerating the first 1000 minimal trap spaces.}
	\label{fig:BBM_cumul}
\end{figure}

@@

	\addplot[
		smooth,
		thick,
		%dashed,
		color=teal,
		mark=square,
	]
	coordinates {
		(0.5, 164)(1, 190)(2, 214)(5, 249)(10, 266)(20, 291)(50, 317)(100, 328)(180, 338)
	}; \addlegendentry{\texttt{maxSAT}}

	\addplot[
		smooth,
		thick,
		%dashed,
		color=green,
		mark=star,
	]
	coordinates {
		(0.5, 143)(1, 176)(2, 195)(5, 222)(10, 249)(20, 264)(50, 287)(100, 308)(180, 322)
	}; \addlegendentry{\texttt{CP}}

	\addplot[
		smooth,
		thick,
		dashed,
		color=magenta,
		mark=star,
	]
	coordinates {
		(0.5, 144)(1, 177)(2, 196)(5, 223)(10, 250)(20, 265)(50, 288)(100, 309)(180, 323)
	}; \addlegendentry{\texttt{ILP}}

	\addplot[
		smooth,
		thick,
		dashed,
		color=violet,
		mark=star,
	]
	coordinates {
		(0.5, 179)(1, 211)(2, 244)(5, 274)(10, 304)(20, 327)(50, 339)(100, 347)(180, 349)
	}; \addlegendentry{\texttt{ASP}}

	\end{semilogxaxis}
	\end{tikzpicture}
	\caption{Cumulative numbers of random models solved by the competing methods with respect to enumerating all the minimal trap spaces.}
	\label{fig:random_cumul_first}
\end{figure}

@@

%Table~\ref{tab:result_N_K} shows the experimental results on N-K models.
%Column \(n\) denotes the network size.
%Columns \texttt{LQM} and \texttt{PBN} show the results of \texttt{bioLQM} and \texttt{PyBoolNet}, respectively.
%For each method, the number outside the parentheses indicates the number of failures, whereas the number inside the parentheses indicates the average running time (in seconds).
%Note that when computing the average running time, if the running time exceeds 180s, it is considered as 180s.
%From these results, we obtained several observations consistent with those obtained for real-world models.

Figure~\ref{fig:random_cumul_first} shows cumulative numbers of random models solved by the competing methods with respect to enumerating all the minimal trap spaces.
The number of succeeded models within three minutes for each method is as follows: \texttt{bioLQM} (0), \texttt{PyBoolNet} (320), \texttt{mpbn} (0), \texttt{Trappist-maxSAT} (338), \texttt{Trappist-CP} (0), \texttt{Trappist-ILP} (0), \texttt{Trappist-ASP} (349).
We can see that \texttt{Trappist-ASP} is the method that could handle the most number of models.
Note that none of the other methods could handle the model failed by \texttt{Trappist-ASP}.
Upon closer inspection, we obtained several observations consistent with those obtained for real-world models as follows.

@@

Second, surprisingly \texttt{bioLQM} could not handle any model.
One of the reason may be that the BDD characterizing all trap spaces is too large, and its computation is slow.
In addition, having too many generic trap spaces before the filtering process may be also a reason.
It is apparent because the network size is large (\(\geq 100\)) and the Boolean functions are not simple.
